Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-04-23 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
some changes
changes

%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
% General:
The detection of gravitational waves produced by coalescing compact binaries
allows us to explore the dynamical, nonlinear regime of general relativity
and to constrain modifications to it.
%
Some of these gravitational-wave events have sufficiently high
signal-to-noise ratio in the merger, allowing us to probe the relaxation of the remnant black
hole to its final, stationary state --- the so-called black-hole ringdown, which is
characterized by a set of quasinormal modes.
% Main question:
Can we use the ringdown to constrain deviations to general relativity,
as predicted by several of its contenders?
% What we do:
Here, we  address this question by using an inspiral-merger-ringdown waveform
model in the effective-one-body formalism, augmented with a parametrization of
the ringdown based on an expansion in the final black hole's spin.
%
We give a prescription on how to include in this waveform model the quasinormal mode frequencies
calculated on a theory-by-theory basis. In particular, we focus on theories that deviate
from general relativity by higher-order curvature corrections, namely,
Einstein-dilaton-Gauss-Bonnet, dynamical Chern-Simons, cubic- and quartic-order
effective-field-theories of general relativity.
%
% Main conclusion:
We confront the resultant waveform models against the two loudest
gravitational-wave ringdown signals observed so far, notably, GW150914 and GW200129.
%
We find that while the Einstein-dilaton-Gauss-Bonnet theory cannot be
constrained with these events, we can place upper bounds on the coupling
constants of cubic- ($\ell_{\rm cEFT} \leqslant 38.2$~km) and quartic-order
($\ell_{\rm qEFT} \leqslant 51.3$~km) effective-field-theories of general
relativity, and of dynamical Chern-Simons gravity ($\ell_{\rm dCS} \leqslant 38.7$~km).
%
The latter result is a concrete example of a theory presently unconstrained by
inspiral-only analyses which, however, can be constrained by merger-ringdown
studies with current gravitational-wave data.
%
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%

@@

\ab{We will treat the modified theories of gravity as EFTs, and focus on 
finite-size effects (see, e.g., Ref.~\cite{Sennett:2019bpc}). Thus, for each gravity theory the fundamental lenght-scale 
$\ell \lesssim GM/c^2$, where $M$ is the mass of the BH. This implies 
that observable deviations from GR present in those theories arise from modifications 
to the Kerr geometry of each individual BH. Those finite-size effects can manifest themselves in 
the QNMs of the remnant produced by the merger, but also in the GW phasing of the inspiral through 
corrections to the GR spin-induced quadrupole and Love numbers. However, here we will not consider the latter, 
instead, we will only study the impact of finite-size effects on the QNMs of the 
remnant.}

@@

Our last example of modified gravity theories are the so-called effective field
theories (EFT) of GR~\cite{Endlich:2017tqa,Sennett:2019bpc,deRham:2020ejn,Cano:2020cao,Cano:2021myl}.
%
They are described by the action
%
\begin{equation} \label{eq:action_eft}
    S_{\rm EFT} = \frac{1}{16 \pi}
    \int \dd^4x \sqrt{-g}
    \left[ R
    +
    \sum_{n \geqslant 2} \ell_{\rm EFT}^{2n - 2}\, L^{(2n)}
    \right]\,,
\end{equation}
%
where $\ell_{\rm EFT}$ is a length-scale assumed to be small compared to the length-scale $M$
associated with a BH (i.e., $\ell_{\rm EFT} / M \ll 1$), and $L^{(2n)}$ are corrections
that introduce higher-order curvature tensors (with $2n$ metric derivatives).

@@

Having reviewed the modified gravity theories that we will consider, we now present
the sequential building blocks for the waveform model we use to test these theories
against GW observations.
%
We start by reviewing the parametrized ringdown spin-expansion coefficients
(ParSpec) framework~\cite{Maselli:2019mjd} in Sec.~\ref{sec:review_parspec}.
%
Next, in Sec.~\ref{sec:review_pSEOB}, we review our baseline
parametrized inspiral(--plunge)-merger-ringdown (IMR) waveform
model~\cite{Brito:2018rfr,Ghosh:2021mrv}, and explain how we extend
it to include the ParSpec.
%
Finally, in Sec.~\ref{sec:theory_specific_qnm}, we show how we can map
theory-specific QNM calculations in modified gravity theories onto the
free coefficients in the ParSpec framework.
%
Ultimately, this provides us with an IMR waveform model, with the ringdown
portion of the model informed by QNM calculations in specific beyond-GR
theories.

@@

Finally, we remark that in the GR limit ($\gamma \to 0$)
the series~\eqref{eq:kerr_expansion} truncated at $N_{\rm max} = 4$, reproduces with 1\% accuracy
the GR QNMs for BH's spins $\chi \lesssim 0.7$.
%
The values of the fitting coefficients $\omega_{\ell m n}^{(j)}$ and $\tau_{\ell m n}^{(j)}$
can be found in Ref.~\cite{Maselli:2019mjd}.
%
In Ref.~\cite{Carullo:2021dui}, the fitting coefficients were calculated up to $N_{\rm max} = 9$,
which extend the validity of the spin-expansion up to $\chi \lesssim 0.99$.
%
As we will discuss in Sec.~\ref{sec:pe}, the expansion to $N_{\rm max} = 4$ is
sufficient for our purposes. For convenience we list the coefficients in the case of
GR in Table~\ref{tab:ref_theories_qnms}.

@@

For the polarizations of the GW signal (in the observer's frame) we have
%
\begin{equation}
h_+(\iota,\varphi_0;t ) - i h_\times(\iota,\varphi_0;t) = \sum_{l, m} {}_{-\!2}Y_{l m}(\iota,\varphi_0)\, h_{l m}(t),
\end{equation}
%
where ${}_{-\!2}Y_{l m}(\iota,\varphi_0)$ are the $-2$ spin-weighted spherical
harmonics.
%
% --------------------------------
% HS: maybe not necessary anymore:
% \footnote{Note that we use $l$ to denote the angular dependence index and $\ell$ to denote the coupling constant in Sec.~\ref{sec:review_parspec}.}
% --------------------------------
%
As our baseline model, that is, the GR model upon which non-GR modifications
are added, we use the computationally efficient (time-domain) multipolar
waveform model for quasicircular spin-aligned BBHs described in
Ref.~\cite{Mihaylov:2021bpf}, which contains the modes, $(l, |m|)=(2,2),(2,1)$, $(3,3)$, $(4,4)$,
and $(5,5)$. Such a model was built by applying the post-adiabatic approximation~\cite{Nagar:2018gnk}
to the multipolar spin-aligned effective-one-body (EOB) waveform model of Refs.~\cite{Bohe:2016gbl,Cotesta:2018fcv} (henceforth we refer to our baseline model as \SEOB{}\footnote{This waveform
model is available in \texttt{LALSuite}~\cite{lalsuite} as the \texttt{SEOBNRv4HM\_PA}
waveform approximant.}).

@@

\subsection{Interpretation of our results}
\label{sec:remarks}

There are two conditions that we must verify before we can confidently claim to
have placed a constraint on $\ell_{\rm th}$. First, as we have explained in Sec.~\ref{sec:review_theories}, all theories that we
consider must be interpreted as an effective field theory, meaning that it should be
considered valid only below an energy scale, or equivalently, a length-scale.
%
As a cut-off length-scale for the validity of the EFT we use,
%
\begin{equation}
\Lambda_{\rm EFT} (\varepsilon, \mathfrak{m}) = \varepsilon \, \frac{G \mathfrak{m}}{c^2} \,,
\label{eq:def_cutoff}
\end{equation}
%
where $\varepsilon$ is a dimensionless number and $\mathfrak{m}$ is the median
value of one of the mass scales involved in the problem.
%
We note that $\Lambda_{\rm EFT}$ has dimensions of length and hence can be compared to the coupling strength $\ell_{\rm th}$.
%
Here we explore the range $\varepsilon \in [0, 1]$, but following
the Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr} we quote our final results using
$\varepsilon = 1/2$.
% %
% In Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}, $\varepsilon = 1/2$ was used,
% but here we explore $\varepsilon \in [0, 1]$ for generality.
% %
We say that a bound has been placed on $\ell_{\rm th}$, if most of the PDF $P(\ell_{\rm th} | d)$ support is in the interval
$[0, \, \Lambda_{\rm EFT}(1/2, \mathfrak{m})]$.
%
In practice, this can be quantified through the cumulative distribution function
(CDF) associated with the marginalized posterior distribution $P(\ell_{\rm th} | d)$, namely
%
\begin{equation}
P(\ell_{\rm th} \leqslant \ell^{\rm max}_{\rm th} | d) = \int_{0}^{\,\ell^{\rm max}_{\rm th}} \dd \ell'\, P(\ell' | d).
\label{eq:def_cdf}
\end{equation}
%
For instance, we require that for a bound at 90\% credible level to be placed on $\ell_{\rm th}$ that
%
\begin{equation}
% \textrm{CDF}(\ell) \leqslant \textrm{CDF}(T_{\rm EFT}(\varepsilon; \, \mathfrak{m})).
P(\ell_{\rm th} \leqslant \Lambda_{\rm EFT} | d) \leqslant 0.9,
\quad \textrm{(EFT bound)},
\label{eq:eft_bound}
\end{equation}
%
where we let $\ell^{\rm max}_{\rm th} = \Lambda_{\rm EFT}$ in Eq.~\eqref{eq:def_cdf}, and
likewise for other credibility percentiles.

@@

\begin{figure}[t]
\includegraphics[width=\columnwidth]{figs/dcs_cdf_varying_threshold.pdf}
\caption{(Color online).~Similar to Fig.~\ref{fig:edgb_cdf}, but for dCS gravity.
% The cumulative distribution function evaluated at the
% cut-off $\Lambda_{\rm EFT}(\varepsilon, \gm)$ as a function of the parameter $\varepsilon$ for both
% $\gm = m_2$ (solid and dashed lines) and $\gm = M_{f}$ (dot-dashed and dotted lines) without
% (blue lines) and with (orange lines) linear-in-spin QNM corrections.
% The horizontal lines lay at 0.68 and 0.90.
%
We see that the CDF curves for $\gm = M_{f}$ are above 90\% for $\varepsilon =
1/2$ for both events, with and without including the $n = 1$ dCS corrections to
the dominant QNM.
}
\label{fig:dcs_cdf}
\end{figure}

We now consider dCS gravity where the main results are summarized in Fig.~\ref{fig:dCS_exec_sum}.
%
As with EdGB gravity (see Sec.~\ref{sec:results_edgb}) although the PDF of $\ell_{\rm dCS}$
is peaked away from 0, this does not signify a deviation from GR, as we have verified that,
%
\begin{equation}
    \gamma_{\rm CS} = \left( \frac{c^2 \ell_{\rm dCS}}{G M_{f}} \right)^{4}\,,
\end{equation}
%
does indeed peak at zero indicating consistency with GR, similarly to what is
shown in Fig.~\ref{fig:gamma_gb} for $\gamma_{\rm EdGB}$.
%
We also see that in both cases the inclusion of leading-order--in-spin
correction to the QNM displaces the posteriors towards smaller values of
$\ell_{\rm dCS}$. This can be seen more evidently by looking at the location of
posterior peaks.
%
Finally, in the bottom panel, we show the combined result for both events.
% \agcomm{Just a thought.  Since, Figs 3,6,8,10 and Figs.4,6,9,11 show the same basic thing,
% should we have one description at the top of the section for each figure? We can follow this up
% by what is special for each theory in each section? Currently, we are repeating a lot of the
% language to describe these figures for each theory?}

@@

\begin{figure}[t]
\includegraphics[width=0.9\columnwidth]{figs/tmp_GW150914_intrinsic_params_remnant.pdf}
\caption{(Color online).~Corner plot showing that the inferred final spin $\chi_f$,
remnant mass $M_{\rm f}$ for GW150914, using the same waveform model,
but without (purple contours) and with the non-GR parameters
different from zero (blue contours), for
dCS gravity and $N_{\rm max}=0$.
%
The contours represent 90\% credible levels.
%
We see that the introduction of the non-GR parameters does not bias the
inference on the source parameters as required by the ParSpec.
}
\label{fig:corner_plot}
\end{figure}

@@

\hscomm{I moved the plot here as we discussed.}
%
As an additional check, to verify the robustness of our constraint, we show in
Fig.~\ref{fig:corner_plot}, the final spin $\chi_{\rm f}$ and remnant mass
$M_{\rm f}$ for GW150914 for GR and dCS gravity.
%
We see that our \pSEOB~waveform model does not introduce substantial changes
to the GR estimates on these parameters, as required by the ParSpec expansion
(see discussion in Sec.~\ref{sec:review_parspec}).
%
In fact, we observed no bias on the estimation of $\Mf$ and $\chi_{\rm f}$
for all theories considered here.
%
In Appendix~\ref{app:big_contour} we show how all other intrinsic parameter parameters
remain unbiased.

@@

Let us now consider the quartic EFT of GR, as our final example.
%
In Fig.~\ref{fig:qEFT_exec_sum} we show the posteriors on $\ell_{\rm qEFT}$ for GW150914 (top panel),
GW200129 (middle panel) for $N_{\rm max} = 0$, which are qualitatively similar to the cubic EFT of GR.
%
We find that both the EFT~\eqref{eq:eft_bound} and ParSpec~\eqref{eq:parspec_bound} are satisfied,
hence allowing us to infer bounds on $\ell_{\rm qEFT}$.
%
As Fig.~\ref{fig:qEFT_cdf} shows the CDFs for both events reach 90\% for $\gm = M_{\rm f}$ before $\varepsilon = 1/2$.
%
Consequently, from the joint posterior of $\ell_{\rm qEFT}$ we quote our final result,
%
\begin{equation}
    \ell_{\rm qEFT} \leqslant 51.3~\textrm{km} \quad \textrm{at 90\% credible level}\,.
\end{equation}
%
% \abcomm{Please update the paragraph below.}
In this theory, we have considered only $N_{\rm max} = 0$.
%
\hs{We found that by adding the spin corrections and for $\ell_{\rm qEFT} \gtrsim 80$~km, resulted in waveforms which
had a ringdown segment larger (sometimes seconds long) that the inspiral-plunge segment in the detectors frequency bands,
making the parameter estimation challenging.}
% for which the ringdown portions of the larger than the inspiral-plunge part. \abcomm{The last sentence
% has some problems in English. I do not udnerstand it.}
%
To overcome this issue we have lowered the value of $\ell^{\rm max}_{\rm qEFT}$, but by doing so we have obtained posteriors
which were flat, just as our prior, and were thus uninformative. Hence, we do not quote any results for $N_{\rm max} = 1$.
%
Table~\ref{tab:summary_qeft} summarizes our findings for the quartic EFT of GR.
%
\begin{figure}[t]
\includegraphics[width=\columnwidth]{figs/qeft_posteriors_combined.pdf}
\caption{(Color online).~Posterior distribution function for the coupling constant $\ell_{\rm qEFT}$ for
GW150914 (top panel), GW200129 (middle panel) and combined events (bottom panel).
}
\label{fig:qEFT_exec_sum}
\end{figure}

@@

\begin{table}[h]
\begin{tabular}{l l c c c}
\hline
\hline
$N_{\rm max}$ & Event & EFT    & ParSpec & Constraint ($\gm = M_{f}$) \\
              &       & bound? & bound?  &                            \\
\hline
  & GW150914 & Yes & Yes  & $\ell_{\rm qEFT} \leqslant 51.7$~km \\
0 & GW200129 & Yes & Yes  & $\ell_{\rm qEFT} \leqslant 54.8$~km \\
  & Combined &     &      & \cellcolor{black!10}$\ell_{\rm qEFT} \leqslant 51.3$~km \\
% \hline
%   & GW150914 &  &     &    \\
% 1 & GW200129 &  &     &    \\
%   & Combined &  &     &    \\
\hline
\hline
\end{tabular}
\caption{Detailed summary of our results the quartic EFT of GR for GW150914, GW200129, and
combined events using $\gm = M_{f}$, $\varepsilon = 1/2$, and $N_{\rm max} = 0$. The quoted
result correspond to 90\% credible values.
}
\label{tab:summary_qeft}
\end{table}

@@

We presented an unified framework that combines the ParSpec framework
to model deviations to the GR QNMs~\cite{Maselli:2019mjd} with the \pSEOB~waveform model~\cite{Brito:2018rfr,Ghosh:2021mrv}.
%
We showed with concrete examples, how theory-specific QNM calculations of
slowly-rotating BHs in modified gravity theories can be mapped onto the
non-GR parameters of the ParSpec formalism.
%
The resulting \pSEOB~waveform model not bias (relative to GR) the inference of
the intrinsic binary parameters as required by ParSpec (see, in particular, Fig.~\ref{fig:corner_plot} and
Appendix~\ref{app:big_contour},~Fig.~\ref{fig:corner_plot_all}).
%
Put together this allowed us to test four modified gravity theories (EdGB, dCS,
cubic, and quartic EFTs of GR) using observational data from the LVK events
GW150914 and GW200129.

We found, that within the interpretation these theories as EFTs and the
region of validity of the ParSpec framework, that the coupling constant of dCS gravity is bound as
$\ell_{\rm dCS} \leqslant 34.5$~km, at 90\% credible level, when stacking the posteriors
of GW150914 and GW200129. This is the first GW-alone constrain on this theory.
%
In contrast, we could not place any bounds on the coupling constant of EdGB
gravity $\ell_{\rm EdGB}$.

This dichotomy between the two theories has a counterpart with works that considered
the inspiral part of the GW signal alone~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}.
%
In those cases, it is dCS gravity which cannot be constrained with current GW
events, while EdGB can.
%
We emphasize that both theories (and the cubic EFT of GR also studied here) all predict the same exponent $p$ in ParSpec.
%
Hence, our results show how the inclusion of theory-specific information into the ParSpec framework
can result in different outcomes for different theories, even if they predict same value of $p$.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
% General:
The detection of gravitational waves produced by coalescing compact binaries
allows us to explore the dynamical, nonlinear regime of general relativity
and to constrain modifications to it.
%
Some of these gravitational-wave events have sufficiently high
signal-to-noise ratio in the merger, allowing us to probe the relaxation of the remnant black
hole to its final, stationary state --- the so-called black-hole ringdown, which is
characterized by a set of quasinormal modes.
% Main question:
Can we use the ringdown to constrain deviations to general relativity,
as predicted by several of its contenders?
% What we do:
Here, we  address this question by using an inspiral-merger-ringdown waveform
model in the effective-one-body formalism, augmented with a parametrization of
the ringdown based on an expansion in the final black hole's spin.
%
We give a prescription on how to include in this waveform model the quasinormal mode frequencies
calculated on a theory-by-theory basis. In particular, we focus on theories that deviate
from general relativity by higher-order curvature corrections, namely,
Einstein-dilaton-Gauss-Bonnet, dynamical Chern-Simons, cubic- and quartic-order
effective-field-theories of general relativity.
%
% Main conclusion:
We confront the resultant waveform models against the two loudest
gravitational-wave ringdown signals observed so far, notably, GW150914 and GW200129.
%
We find that while the Einstein-dilaton-Gauss-Bonnet theory cannot be
constrained with these events, we can place upper bounds on the \sout{coupling
constants} \ab{fundamental length-scale} of cubic- ($\ell_{\rm cEFT} \leqslant 38.2$~km) and quartic-order
($\ell_{\rm qEFT} \leqslant 51.3$~km) effective-field-theories of general
relativity, and of dynamical Chern-Simons gravity ($\ell_{\rm dCS} \leqslant 38.7$~km).
%
The latter result is a concrete example of a theory presently unconstrained by
inspiral-only analyses which, however, can be constrained by merger-ringdown
studies with current gravitational-wave data.
%
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%

@@

\ab{We will treat the modified theories of gravity as EFTs, and focus on 
finite-size effects (see, e.g., Ref.~\cite{Sennett:2019bpc}). Thus, for each gravity theory we impose that 
the fundamental length-scale $\ell \lesssim GM/c^2$, where $M$ is the mass of the BH. This implies 
that observable deviations from GR present in those theories arise from modifications 
to the Kerr geometry of each individual BH. Those finite-size effects can manifest themselves in 
the QNMs of the remnant produced by the merger, but also in the GW phasing of the inspiral through 
corrections to the GR spin-induced quadrupole and Love numbers. However, here we will not consider the latter, 
instead, we will only study the impact of finite-size effects on the QNMs of the 
remnant.}

@@

Our last example of modified gravity theories are the so-called 
EFTs of GR~\cite{Endlich:2017tqa,Sennett:2019bpc,deRham:2020ejn,Cano:2020cao,Cano:2021myl}.
%
They are described by the action
%
\begin{equation} \label{eq:action_eft}
    S_{\rm EFT} = \frac{1}{16 \pi}
    \int \dd^4x \sqrt{-g}
    \left[ R
    +
    \sum_{n \geqslant 2} \ell_{\rm EFT}^{2n - 2}\, L^{(2n)}
    \right]\,,
\end{equation}
%
where $\ell_{\rm EFT}$ is a length-scale assumed to be small compared to the length-scale $M$
associated with a BH (i.e., $\ell_{\rm EFT} / M \ll 1$), and $L^{(2n)}$ are corrections
that introduce higher-order curvature tensors (with $2n$ metric derivatives).

@@

Having reviewed the modified gravity theories that we will consider, we now present
the sequential building blocks for the waveform model we use to test these theories
against GW observations.
%
We start by reviewing the parametrized ringdown spin-expansion coefficients
(ParSpec) framework~\cite{Maselli:2019mjd} in Sec.~\ref{sec:review_parspec}.
%
Next, in Sec.~\ref{sec:review_pSEOB}, we review our baseline
parametrized IMR waveform model~\cite{Brito:2018rfr,Ghosh:2021mrv}, and explain how we extend
it to include the ParSpec.
%
Finally, in Sec.~\ref{sec:theory_specific_qnm}, we show how we can map
theory-specific QNM calculations in modified gravity theories onto the
free coefficients in the ParSpec framework.
%
Ultimately, this provides us with an IMR waveform model, with the ringdown
portion of the model informed by QNM calculations in specific beyond-GR
theories.

@@

Finally, we remark that in the GR limit ($\gamma \to 0$)
the series~\eqref{eq:kerr_expansion} truncated at $N_{\rm max} = 4$, reproduces with 1\% accuracy
the GR QNMs for BH's spins $\chi_{\rm f} \lesssim 0.7$.
%
The values of the fitting coefficients $\omega_{\ell m n}^{(j)}$ and $\tau_{\ell m n}^{(j)}$
can be found in Ref.~\cite{Maselli:2019mjd}.
%
In Ref.~\cite{Carullo:2021dui}, the fitting coefficients were calculated up to $N_{\rm max} = 9$,
which extend the validity of the spin-expansion up to $\chi_{\rm f} \lesssim 0.99$.
%
As we will discuss in Sec.~\ref{sec:pe}, the expansion to $N_{\rm max} = 4$ is
sufficient for our purposes. For convenience we list the coefficients in the case of
GR in Table~\ref{tab:ref_theories_qnms}.

@@

For the polarizations of the GW signal (in the observer's frame) we have
%
\begin{equation}
h_+(\iota,\varphi_0;t ) - i h_\times(\iota,\varphi_0;t) = \sum_{l, m} {}_{-\!2}Y_{l m}(\iota,\varphi_0)\, h_{l m}(t),
\end{equation}
%
where ${}_{-\!2}Y_{l m}(\iota,\varphi_0)$ are the $-2$ spin-weighted spherical
harmonics.
%
% --------------------------------
% HS: maybe not necessary anymore:
% \footnote{Note that we use $l$ to denote the angular dependence index and $\ell$ to denote the coupling constant in Sec.~\ref{sec:review_parspec}.}
% --------------------------------
%
As our baseline model, that is, the GR model upon which non-GR modifications
are added, we use the computationally efficient (time-domain) multipolar
waveform model for quasicircular spin-aligned BBHs described in
Ref.~\cite{Mihaylov:2021bpf}, which contains the modes, $(l, |m|)=(2,2),(2,1)$, $(3,3)$, $(4,4)$,
and $(5,5)$. Such a model was built by applying the post-adiabatic approximation~\cite{Nagar:2018gnk}
to the multipolar spin-aligned EOB waveform model of Refs.~\cite{Bohe:2016gbl,Cotesta:2018fcv} (henceforth we refer to our baseline model as \SEOB{}\footnote{This waveform
model is available in \texttt{LALSuite}~\cite{lalsuite} as the \texttt{SEOBNRv4HM\_PA}
waveform approximant.}).

@@

\subsection{EFT interpretation of our results}
\label{sec:remarks}

There are two conditions that we must verify before we can confidently claim to
have placed a constraint on $\ell_{\rm th}$. First, as we have explained in Sec.~\ref{sec:review_theories}, all theories that we
consider must be interpreted as an EFT, meaning that it should be
considered valid only below an energy scale, or equivalently, a length-scale.
%
As a cut-off length-scale for the validity of the EFT we use,
%
\begin{equation}
\Lambda_{\rm EFT} (\varepsilon, \mathfrak{m}) = \varepsilon \, \frac{G \mathfrak{m}}{c^2} \,,
\label{eq:def_cutoff}
\end{equation}
%
where $\varepsilon$ is a dimensionless number and $\mathfrak{m}$ is the median
value of one of the mass scales involved in the problem.
%
We note that $\Lambda_{\rm EFT}$ has dimensions of length and hence can be compared to the coupling strength $\ell_{\rm th}$.
%
Here we explore the range $\varepsilon \in [0, 1]$, but following
the Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr} we quote our final results using
$\varepsilon = 1/2$.
% %
% In Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}, $\varepsilon = 1/2$ was used,
% but here we explore $\varepsilon \in [0, 1]$ for generality.
% %
We say that a bound has been placed on $\ell_{\rm th}$, if most of the PDF $P(\ell_{\rm th} | d)$ support is in the interval
$[0, \, \Lambda_{\rm EFT}(1/2, \mathfrak{m})]$.
%
In practice, this can be quantified through the cumulative distribution function
(CDF) associated with the marginalized posterior distribution $P(\ell_{\rm th} | d)$, namely
%
\begin{equation}
P(\ell_{\rm th} \leqslant \ell^{\rm max}_{\rm th} | d) = \int_{0}^{\,\ell^{\rm max}_{\rm th}} \dd \ell'\, P(\ell' | d).
\label{eq:def_cdf}
\end{equation}
%
For instance, we require that for a bound at 90\% credible level to be placed on $\ell_{\rm th}$ that
%
\begin{equation}
% \textrm{CDF}(\ell) \leqslant \textrm{CDF}(T_{\rm EFT}(\varepsilon; \, \mathfrak{m})).
P(\ell_{\rm th} \leqslant \Lambda_{\rm EFT} | d) \leqslant 0.9,
\quad \textrm{(EFT bound)},
\label{eq:eft_bound}
\end{equation}
%
where we let $\ell^{\rm max}_{\rm th} = \Lambda_{\rm EFT}$ in Eq.~\eqref{eq:def_cdf}, and
likewise for other credibility percentiles.

@@

\begin{figure}[t]
\includegraphics[width=\columnwidth]{figs/dcs_cdf_varying_threshold.pdf}
\caption{(Color online).~Similar to Fig.~\ref{fig:edgb_cdf}, but for dCS gravity.
% The cumulative distribution function evaluated at the
% cut-off $\Lambda_{\rm EFT}(\varepsilon, \gm)$ as a function of the parameter $\varepsilon$ for both
% $\gm = m_2$ (solid and dashed lines) and $\gm = M_{f}$ (dot-dashed and dotted lines) without
% (blue lines) and with (orange lines) linear-in-spin QNM corrections.
% The horizontal lines lay at 0.68 and 0.90.
%
We see that the CDF curves for $\gm = M_{\rm f}$ are above 90\% for $\varepsilon =
1/2$ for both events, with and without including the $n = 1$ dCS corrections to
the dominant QNM.
}
\label{fig:dcs_cdf}
\end{figure}

We now consider dCS gravity where the main results are summarized in Fig.~\ref{fig:dCS_exec_sum}.
%
As with EdGB gravity (see Sec.~\ref{sec:results_edgb}) although the PDF of $\ell_{\rm dCS}$
is peaked away from 0, this does not signify a deviation from GR, as we have verified that,
%
\begin{equation}
    \gamma_{\rm CS} = \left( \frac{c^2 \ell_{\rm dCS}}{G M_{\rm f}} \right)^{4}\,,
\end{equation}
%
does indeed peak at zero indicating consistency with GR, similarly to what is
shown in Fig.~\ref{fig:gamma_gb} for $\gamma_{\rm EdGB}$.
%
We also see that in both cases the inclusion of leading-order--in-spin
correction to the QNM displaces the posteriors towards smaller values of
$\ell_{\rm dCS}$. This can be seen more evidently by looking at the location of
posterior peaks.
%
Finally, in the bottom panel, we show the combined result for both events.
% \agcomm{Just a thought.  Since, Figs 3,6,8,10 and Figs.4,6,9,11 show the same basic thing,
% should we have one description at the top of the section for each figure? We can follow this up
% by what is special for each theory in each section? Currently, we are repeating a lot of the
% language to describe these figures for each theory?}

@@

\begin{figure}[t]
\includegraphics[width=0.9\columnwidth]{figs/tmp_GW150914_intrinsic_params_remnant.pdf}
\caption{(Color online).~Corner plot showing that the inferred final spin $\chi_{\rm f}$,
remnant mass $M_{\rm f}$ for GW150914, using the same waveform model,
but without (purple contours) and with the non-GR parameters
different from zero (blue contours), for
dCS gravity and $N_{\rm max}=0$.
%
The contours represent 90\% credible levels.
%
We see that the introduction of the non-GR parameters does not bias the
inference on the source parameters as required by the ParSpec.
}
\label{fig:corner_plot}
\end{figure}

@@

%
As an additional check, to verify the robustness of our constraint, we show in
Fig.~\ref{fig:corner_plot}, the final spin $\chi_{\rm f}$ and remnant mass
$M_{\rm f}$ for GW150914 for GR and dCS gravity.
%
We see that our \pSEOB~waveform model does not introduce substantial changes
to the GR estimates on these parameters, as required by the ParSpec expansion
(see discussion in Sec.~\ref{sec:review_parspec}).
%
In fact, we observed no bias on the estimation of $\Mf$ and $\chi_{\rm f}$
for all theories considered here.
%
\ab{For completness,} in Appendix~\ref{app:big_contour} we also 
show how all other intrinsic parameter parameters remain unbiased.

@@

Let us now consider the quartic EFT of GR, as our final example.
%
In Fig.~\ref{fig:qEFT_exec_sum} we show the posteriors on $\ell_{\rm qEFT}$ for GW150914 (top panel),
GW200129 (middle panel) for $N_{\rm max} = 0$, which are qualitatively similar to the cubic EFT of GR.
%
We find that both the EFT~\eqref{eq:eft_bound} and ParSpec~\eqref{eq:parspec_bound} are satisfied,
hence allowing us to infer bounds on $\ell_{\rm qEFT}$.
%
As Fig.~\ref{fig:qEFT_cdf} shows the CDFs for both events reach 90\% for $\gm = M_{\rm f}$ before $\varepsilon = 1/2$.
%
Consequently, from the joint posterior of $\ell_{\rm qEFT}$ we quote our final result,
%
\begin{equation}
    \ell_{\rm qEFT} \leqslant 51.3~\textrm{km} \quad \textrm{at 90\% credible level}\,.
\end{equation}
%
% \abcomm{Please update the paragraph below.}
In this theory, we have considered only $N_{\rm max} = 0$.
%
\ab{Indeed, we find that when adding the spin corrections and for $\ell_{\rm qEFT} \gtrsim 80$~km, 
the waveforms can have a ringdown segment larger (sometimes seconds long) than the inspiral-plunge segment in the detectors frequency band, making the parameter estimation challenging.} \abcomm{The reference to 80 km comes out of the blue. Doesn't it? Why did you refer to it?}
% for which the ringdown portions of the larger than the inspiral-plunge part. \abcomm{The last sentence
% has some problems in English. I do not udnerstand it.}
%
To overcome this issue we have lowered the value of $\ell^{\rm max}_{\rm qEFT}$, but by doing so we have obtained posteriors
which were flat, just as our prior, and were thus uninformative. Hence, we do not quote any results for $N_{\rm max} = 1$.
%
Table~\ref{tab:summary_qeft} summarizes our findings for the quartic EFT of GR.
%
\begin{figure}[t]
\includegraphics[width=\columnwidth]{figs/qeft_posteriors_combined.pdf}
\caption{(Color online).~Posterior distribution function for the coupling constant $\ell_{\rm qEFT}$ for
GW150914 (top panel), GW200129 (middle panel) and combined events (bottom panel).
}
\label{fig:qEFT_exec_sum}
\end{figure}

@@

\begin{table}[h]
\begin{tabular}{l l c c c}
\hline
\hline
$N_{\rm max}$ & Event & EFT    & ParSpec & Constraint ($\gm = M_{\rm f}$) \\
              &       & bound? & bound?  &                            \\
\hline
  & GW150914 & Yes & Yes  & $\ell_{\rm qEFT} \leqslant 51.7$~km \\
0 & GW200129 & Yes & Yes  & $\ell_{\rm qEFT} \leqslant 54.8$~km \\
  & Combined &     &      & \cellcolor{black!10}$\ell_{\rm qEFT} \leqslant 51.3$~km \\
% \hline
%   & GW150914 &  &     &    \\
% 1 & GW200129 &  &     &    \\
%   & Combined &  &     &    \\
\hline
\hline
\end{tabular}
\caption{Detailed summary of our results the quartic EFT of GR for GW150914, GW200129, and
combined events using $\gm = M_{\rm f}$, $\varepsilon = 1/2$, and $N_{\rm max} = 0$. The quoted
result correspond to 90\% credible values.
}
\label{tab:summary_qeft}
\end{table}

@@

We presented an unified framework that combines the ParSpec framework
to model deviations to the GR QNMs~\cite{Maselli:2019mjd} with the \pSEOB~waveform model~\cite{Brito:2018rfr,Ghosh:2021mrv}.
%
We showed with concrete examples, how theory-specific QNM calculations of
slowly-rotating BHs in modified gravity theories can be mapped onto the
non-GR parameters of the ParSpec formalism.
%
The resulting \pSEOB~waveform model \ab{does} not bias (relative to GR) the inference of
the intrinsic binary parameters, as required by ParSpec (see, in particular, Fig.~\ref{fig:corner_plot} and
Fig.~\ref{fig:corner_plot_all} in Appendix~\ref{app:big_contour}),
%
Put together this allowed us to test four modified gravity theories (EdGB, dCS,
cubic, and quartic EFTs of GR) using observational data from the LVK events
GW150914 and GW200129. \ab{Our results are summarzied in Table~\ref{tab:bound_summary}.}
